OpenMPIR

### 2017-09-07 paper.tex
Fixed small typo.
Various improvements

@@

When writing task-parallel programs today, one has a large selection of
programming models. Unfortunately, despite having significant overlap in
semantics, parallel programming models like OpenMP, Cilk, Kokkos, HPX, Charm++,
Qthreads, pthreads, MPI, Chapel, UPC, etc. have virtually no ability to
interoperate, both at compile time and run time. In addition, many of these
programming models are implemented as libraries, and therefore have no compiler
support. There has been some recent work on specializing compilers to reason
about libraries \cite{kokkos}, but fragmentation among models presents a
significant challenge for compiler writers, which must choose a single model to
analyze and optimize. An ideal solution would be some some form of common
representation of parallelism that all IRs can share.

Compilers use internal representations (IRs) for implementing analyses and
optimizations. Historically, a major drawback for parallel programs is that IRs
don't have the ability to represent, and therefore reason about, parallelism.
This prevents the application of common optimizations to concurrent code. The
recent work of Schardl et al.  \cite{shardl2017} has shown that real world
compiler IRs can be extended to represent fork-join parallelism. By extending
the LLVM instruction set with three instructions, they were able to capture all
of the semantics to implement the semantics of Cilk, an extension to C. They
suggested that a similar approach could be taken with OpenMP tasks.

@@

\textbf{Internal representations (IRs)} are a crucial part of modern compilers.
By representing code in a generic, hardware-agnostic format, they allow both
multiple frontends for different source languages, and multiple backends for
different hardware, to all take advantage of a series of compiler optimizations
and analyses.

\wmnote{This feels like we're trying to convince someone about LLVM: perhaps
LLVM has gained traction in the academic community for both its simplicity and
relatively painless modification} 
\gsnote{I wanted to give a brief nod to the importance of LLVM, given the venue
(LLVM in HPC). Is this any more satisfactory?} 
Among IRs, LLVM is arguably the most ubiquitous \cite{llvm}. Winner of the
2012 ACM Software System Award, it is in use by frontends ranging from Haskell
to C++, and backends ranging from x86 to PTX \cite{}. Implementing a simple,
extensible \textbf{single static assignment (SSA)} IR, LLVM enables powerful
optimizations that are effective for a wide range of programming languages and
hardware \cite{llvm}.

@@

In this section we discuss a significant challenge in HPC: fragmentation of
parallel programming models. There are many, many, many ways to write
parallel programs \cite{....}. From libraries, to language extensions, to
standalone languages, to combinations of the above, fragmentation of parallel
programming models is a problem that is only getting worse. This means that
when writing tooling, optimizations, or analyses, one has to choose which model
to target. One common difficulty among all of these is that reasoning about
parallel programs is hard \cite{}, and compilers can help \cite{}. This is
exactly analogous the problem LLVM solves: compiler optimizations are
hard to do well, so sharing them is valuable. Tapir is a natural extension

With fragmentation comes questions of compatibility, or composability, of 
different programming models. To illustrate this issue, consider the example of
a program using a library that uses OpenMP to implement parallelism internally. 
If said program uses a different programming model for parallelism, e.g. Cilk
, there is currently no good solution for

Addressing these concerns this is a major motivation for our use of Tapir.
By standardising an IR, in addition to easing the development of optimizations,
one avoids duplicating work across different implementations. We'll return to
this issue later in Section~\ref{Sec:Future} when discussing future work.

@@

\lstset{basicstyle=\tt\small, language=C, 
morekeywords={omp, for, pragma, omp, task, taskwait}}
\begin{figure*}
\begin{center}
\begin{tabular}{| c c c | c c c | c c c |}
\hline
 & 
\begin{lstlisting}
int fib(int n) 
{                    // A
  if (n < 2)
    return n;
  else
  {
    int x, y;
    #pragma omp task
      x = fib(n-1);  // B
    #pragma omp task
      y = fib(n-2);  // C
    #pragma omp taskwait
    return (x+y);    // D
  }
}
\end{lstlisting}
& & &
    \centering{
   \raisebox{-0.6in}{\includegraphics[scale=0.35]{fib-task-graph.pdf}}
    }
& & &
    \centering{
     \raisebox{-0.25in}{\includegraphics[scale=0.35]{fib-schedule.pdf}}
    }
 &  \\
\hline
\end{tabular}
\end{center}
\caption{Code, task graph, and schedule of a simple brute force recursive 
Fibonacci number calculation on two threads.}
\label{Fig:fib-graph-to-schedule}
\end{figure*}

@@

A very important property of the current implementation to discuss is the
backend. This work has \emph{only been on the frontend}, which means we use
the only existing backend for Tapir. The existing backend generates calls into
the Cilk runtime, \texttt{libcilkrts}. This obviously has many implications for
adhering to the OpenMP specification. For example, environment variables such
as \texttt{OMP\_NUM\_THREADS} are ignored, replaced by \texttt{CILK\_NWORKERS}.
We discuss some of these issues below, and return to address others in the
future work section.

There are other issues that had to be overcome to run full OpenMP programs
using Tapir as described above. Generally, in addition the the \texttt{task} 
and \texttt{taskwait} pragmas, a program has to contain initialization OpenMP
pragmas in order to run parallel code. A standard pattern for building
task-parallel OpenMP programs is to insert a \texttt{parallel} pragma to start
the necessary hardware threads, followed immediately by a \texttt{single}
pragma to have only one thread continue on the specified statement. Then the
other threads will get work from spawned tasks, instead of implicitly executing
the same code in a data parallel style. For the purpose of this paper, we took
the shortcut of replacing these pragmas with no-ops. This worked well for all
examples except for one, in which after the \texttt{parallel} and
\texttt{single} pragmas the top level task was called with an unnecessary
\texttt{task} pragma.  Because the wait was implicit and unhandled by our
implementation, the simple fix was to remove the \texttt{task} pragma. This was
the only change required to the source code. Our temporary no-op shortcut, of
course, doesn't follow the OpenMP specification, and should be addressed by
future work.

@@

\begin{figure*}
\begin{tabular}{| c c c | c c c |}
\hline
 & 
\begin{lstlisting}
int fib(int n) 
{                    // A
  if (n < 2)
    return n;
  else
  {
    int x, y;
    #pragma omp task
      x = fib(n-1);  // B
    #pragma omp task
      y = fib(n-2);  // C
    #pragma omp taskwait
    return (x+y);    // D
  }
}
\end{lstlisting}
& & &
\begin{lstlisting}
if.end:                                 
  detach within %syncreg, label %det.achd, label %det.cont

@@

>>>>

\lstset{numbers=left,
         numberstyle=\tiny,
         numbersep=5pt,
         xleftmargin=0.1in
}

@@

When writing task-parallel programs today, one has a large selection of
programming models. Unfortunately, despite having significant overlap in
semantics, parallel programming models like OpenMP, Cilk, Kokkos, HPX, Charm++,
Qthreads, pthreads, MPI, Chapel, UPC, etc. have virtually no ability to
interoperate, both at compile time and run time. In addition, many of these
programming models are implemented as libraries, and therefore have no compiler
support. There has been some recent work on specializing compilers to reason
about parallel programming libraries \cite{Moss_2016}, but fragmentation among
models presents a significant challenge for compiler writers, which must choose
a single model to analyze and optimize. An ideal solution would be some form of
common representation of parallelism in an internal representation \textbf{(IR)}.

Compilers use internal representations (IRs) for implementing analyses and
optimizations. Historically, a major drawback for parallel programs is that IRs
haven't had the ability to represent, and therefore reason about, parallelism.
This has prevented the application of common optimizations to concurrent code.
The recent work of Schardl et al.  \cite{shardl2017} has shown that real world
compiler IRs can be extended to represent fork-join parallelism. By extending
the LLVM instruction set with three instructions, they were able to capture all
of the semantics to implement the semantics of Cilk, an extension to C. They
suggested that a similar approach could be taken with OpenMP tasks.

@@

Internal representations (IRs) are a crucial part of modern compilers.  By
representing code in a generic, hardware-agnostic format, they allow both
multiple frontends for different source languages, and multiple backends for
different hardware, to all take advantage of a series of compiler optimizations
and analyses.

\wmnote{This feels like we're trying to convince someone about LLVM: perhaps
LLVM has gained traction in the academic community for both its simplicity and
relatively painless modification} 
\gsnote{I wanted to give a brief nod to the importance of LLVM, given the venue
(LLVM in HPC). Is this any more satisfactory?} 
Among IRs, LLVM is arguably the most ubiquitous \cite{llvm}. Winner of the
2012 ACM Software System Award, it is in use by frontends ranging from Haskell
to C++, and backends ranging from x86 to PTX \cite{}. Implementing a simple,
extensible single static assignment IR, LLVM enables powerful optimizations
that are effective for a wide range of programming languages and hardware
\cite{llvm}.

@@

In this section we further discuss a significant challenge in HPC:
fragmentation of parallel programming models. There are many, many, many ways
to write parallel programs \cite{....}. From libraries, to language extensions,
to standalone languages, to combinations of the above, fragmentation of
parallel programming models is a problem that is only getting worse. This means
that when writing tooling, optimizations, or analyses, one has to choose which
model to target. One common difficulty among all of these is that reasoning
about parallel programs is hard \cite{}, and compilers can help \cite{}. This
is exactly analogous the problem LLVM solves: compiler optimizations are hard
to do well, so sharing them is valuable. Tapir is a demonstration that the 
the solutions in the case of serial code can be effectively extended to the 
case of parallel code.

With fragmentation comes questions of compatibility, or composability, of 
different programming models. To illustrate this issue, consider the example of
a program using a library that uses OpenMP to implement parallelism internally. 
If said program uses a different programming model for parallelism, e.g. Cilk
, there is currently no good solution for ensuring that the two different backends
will cooperate with management of hardware resources.

Addressing these concerns this is a major motivation for our use of Tapir.
By standardising an IR, in addition to easing the development of optimizations,
one avoids duplicating work across different implementations. It also enables
the possibility of multiple programming models coexisting peacefully. We'll
return to this issue later in Section~\ref{Sec:Discussion} when discussing
future work.

@@

\begin{figure*}
\begin{center}
\begin{tabular}{| c c c | c c c | c c c |}
\hline
 & 
\begin{lstlisting}
int fib(int n) 
{                    // A
  if (n < 2)
    return n;
  else
  {
    int x, y;
    #pragma omp task
      x = fib(n-1);  // B
    #pragma omp task
      y = fib(n-2);  // C
    #pragma omp taskwait
    return (x+y);    // D
  }
}
\end{lstlisting}
& & &
    \centering{
   \raisebox{-0.6in}{\includegraphics[scale=0.35]{fib-task-graph.pdf}}
    }
& & &
    \centering{
     \raisebox{-0.25in}{\includegraphics[scale=0.35]{fib-schedule.pdf}}
    }
 &  \\
\hline
\end{tabular}
\end{center}
\caption{Code, task graph, and schedule of a simple brute force recursive 
Fibonacci number calculation on two threads.}
\label{Fig:fib-graph-to-schedule}
\end{figure*}

@@

This work has \emph{only been on the frontend}, which means we use the only
existing backend for Tapir, which generates code that calls into the Cilk
runtime, \texttt{libcilkrts}. This has implications for adhering to the OpenMP
specification. For example, environment variables such as
\texttt{OMP\_NUM\_THREADS} are ignored, replaced by \texttt{CILK\_NWORKERS}.
We discuss some of these issues below, and return to address others in the
future work section.

There are other issues that had to be overcome to run full OpenMP programs
using Tapir as described above. Generally, in addition the the \texttt{task} 
and \texttt{taskwait} pragmas, a program has to contain initialization OpenMP
pragmas in order to run parallel code. A standard pattern for building
task-parallel OpenMP programs is to insert a \texttt{parallel} pragma to start
the necessary hardware threads, followed immediately by a \texttt{single}
pragma to have only one thread continue on the specified statement. Then the
other threads will get work from spawned tasks, instead of implicitly executing
the same code in a data parallel style. For the purpose of this paper, we took
the shortcut of replacing these pragmas with no-ops. This worked well for all
examples except for one, in which after the \texttt{parallel} and
\texttt{single} pragmas the top level task was called with an unnecessary
\texttt{task} pragma.  Because the wait was implicit and unhandled by our
implementation, the simple fix was to remove the \texttt{task} pragma. This was
the only change required to the source code. Our temporary no-op shortcut, of
course, doesn't follow the OpenMP specification, and should be addressed by
future work. See Section~\ref{Sec:Future} for a further discussion how how the 
implementation can be improved.

@@

\begin{figure*}
\begin{tabular}{| c c c | c c c |}
\hline
 & 
\begin{lstlisting}
int fib(int n) 
{                    // A
  if (n < 2)
    return n;
  else
  {
    int x, y;
    #pragma omp task
      x = fib(n-1);  // B
    #pragma omp task
      y = fib(n-2);  // C
    #pragma omp taskwait
    return (x+y);    // D
  }
}
\end{lstlisting}
& & &
\begin{lstlisting}[language=llvm]
if.end:                                 
  detach within %syncreg, label %det.achd, label %det.cont

@@

The listed Tapir LLVM IR code corresponds to lines 7 through 12 of the C source
code.  Upon entering the \texttt{else} branch, the code immediately detaches
the basic block corresponding to the call to \texttt{fib(n-1)}, labeled
\texttt{det.achd}. The continuation for the first call also immediately
detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
Finally, the continuation for the second call immediately calls sync. This is
actually a nice example of code that will be optimized by Tapir. The LLVM code
shown in Figure~\ref{Fig:Example} is not optimized. With optimizataions turned
on, Tapir will replace a \texttt{detach} of a function call followed
immediately by \texttt{sync} with a simple function call, resulting in faster
code.
